---
title: "Bivariate Correlations and Descriptive Statistics"
author: "Lisa Frueh"
date: '2023-03-07'
output:
  html_document: default
  pdf_document: default
  word_document: default
subtitle: EOH710 Individual Project - Winter 2023
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(spdep)
library(Hmisc)
library(ggcorrplot)
library(plyr)
library(gtsummary)
setwd("~/Library/CloudStorage/OneDrive-DrexelUniversity/Research/EPA STAR/Spatial_Stressors")
```

# Read in Data
Shapefile was created in ArcGIS Pro by joining 2010 NY tracts with tract-level stressors (see Tract_Stressor_Sample.Rmd).
Gal weights file was created in GeoDa.

```{r}
ny <- st_read("./Data/shapefiles_weights/tract_stressor_sample.shp")

#Read in weights file from GeoDa as an nb object--this creates a list of neighbors. In this case, first-order queen contiguity was used to define neighbors.
queen.nb <- read.gal("./Data/shapefiles_weights/tract_ny_queen.gal", region.id=NULL, override.id=TRUE)

#Create spatial weights from neighbors list, style = "W" defines row-standardized weights-sums over all links to n.
lw <- nb2listw(queen.nb, glist=NULL, style="W", zero.policy=TRUE)


```

# Define variables of interest
```{r}
vars <-  c("park_a_per","la_05_10","unemp","fipr_100","poc","relciv_per","disc_youth","statepris1", "unins_18_6", "ambfacs_k")

var_lab <- list(park_a_per = "Park area per 1,000 people", la_05_10 = "Low supermarket access", unemp = "% Unemployed", fipr_100 = "% Under poverty line", poc = "% People of color", relciv_per = "Religious & civic organizations per 1,000 people", disc_youth = "% Disconnected youth", statepris1 = "# Incarcerated in state prisons, per 1,000 people", unins_18_6="% Uninsured (18-64)", ambfacs_k = "Ambulatory medical facilities per 1,000 people")
```

# Pearson rho correlation (non-spatial)
```{r}
#remove geometry
ny_ns <- st_drop_geometry(ny)

#calculate correlation matrix
rho.mat <- round(cor(ny_ns[vars]),2)

#store p values
p.mat <- cor_pmat(ny_ns[vars])

#Visualize
pearson_plot <-
ggcorrplot(rho.mat,
           p.mat = p.mat,
           hc.order=TRUE,
           type = "lower",
           outline.color="white",
           lab=TRUE,
           lab_size = 3,
           insig = "pch",
           pch.col = "gray",
           ggtheme = ggplot2::theme_gray,
           colors = c("#6D9EC1", "white", "#E46726"),
           legend.title = "Pearson r",
           tl.cex = 9
) + 
  scale_x_discrete(labels = c("Uninsured", "Disconnected youth", "Unemployed", "Poverty", "Low food access", "Incarceration", "Parks", "Religious & civic orgs", "Medical facilities")) + 
  scale_y_discrete(labels = c("People of color", "Uninsured", "Disconnected youth", "Unemployed", "Poverty", "Low food access", "Incarceration", "Parks", "Religious & civic orgs")) 

ggsave("pearson_plot.png", pearson_plot, device="png")
pearson_plot

```
# Lee's L statistic
See: Lee (2001). Developing a bivariate spatial association measure: An integration of Pearson's r and Moran's I. J Geograph Syst 3: 369-385

In short, this method deduces a global (and local, but we'll just use global) statistic of bivariate correlation between two spatial variables. 
```{r}

#Check that the Global L statistic is symmetrical, like rho
poc_fipr100 <- lee(ny$poc, ny$fipr_100, lw, length(ny$poc), zero.policy=TRUE, NAOK=TRUE)
poc_fipr100$L

#Monte-carlo simulation for bootstrapped p-value
test<- lee.mc(ny$poc, ny$fipr_100, nsim=99, lw, zero.policy=TRUE, alternative="two.sided", na.action=na.omit)

# It is symmetrical!
fipr100_poc <- lee(ny$fipr_100, ny$poc, lw, length(ny$fipr_100), zero.policy=TRUE, NAOK = TRUE)
fipr100_poc$L

```

## All bivariate pairs
Write a function to run Lee's L-statistic for all combinations of variables specified in 'vars'.
```{r}

#Write a function that computes the L-statistic and p-value (via Monte Carlo simulation) for bivariate association for a pair of variables, x and y, and returns a dataframe.
leepair <- function(d, x, y){
  test <- lee.mc(d[[x]], d[[y]], nsim=99, lw, zero.policy=TRUE, alternative="two.sided", na.action=na.omit)
  L <- test$statistic
  p <- test$p.value
  data.frame(var1=x, var2=y, L=L, p=p)
}

#Apply this function to all possible combinations of variables 1-11 in the list "vars"
#Then, combine these into one dataframe

df_total<-data.frame()
for(x in 1:10){
  for(y in 10:1){
    model <- plyr::ddply(ny, .(), leepair, x=vars[x], y=vars[y])
    df <- data.frame(model)
    df_total <- rbind(df_total,df)
  }
}

#Create a matrix to hold the L values and p values for bivariate combinations
lee.mat <- df_total %>%
  select(-.id, -p) %>%
  spread(var2, L) %>%
  data.frame(., row.names = .$var1) %>%
  select(-var1) %>%
  as.matrix(.)


lee.p.mat <- df_total %>%
  select(-.id, -L) %>%
  spread(var2, p) %>%
  data.frame(., row.names = .$var1) %>%
  select(-var1) %>%
  as.matrix(.)


#Visualize
L_plot <-
ggcorrplot(lee.mat,
           p.mat = lee.p.mat,
           hc.order=TRUE,
           type = "lower",
           outline.color="white",
           lab=TRUE,
           lab_size = 3,
           insig = "pch",
           pch.col = "gray",
           ggtheme = ggplot2::theme_gray,
           colors = c("#6D9EC1", "white", "#E46726"),
           legend.title = "Lee's L",
           tl.cex = 9
) + 
  scale_x_discrete(labels = c("Incarceration", "Parks", "Medical facilities", "Religious & civic orgs", "Unemployed", "Disconnected youth", "Poverty", "People of color", "Uninsured")) + 
  scale_y_discrete(labels = c("Low food access", "Incarceration", "Parks", "Medical facilities", "Religious & civic orgs", "Unemployed", "Disconnected youth", "Poverty", "People of color"))

ggsave("L_plot.png", L_plot, device="png")
L_plot
```

# Descriptive Statistics
```{r}

ny_ns %>%
  select(all_of(vars)) %>%
  tbl_summary(
    label = var_lab,
    type = all_continuous() ~ "continuous2", 
    statistic = all_continuous() ~ c(
      "{mean}",
      "{median} ({p25}, {p75})",
      "{min}, {max}"
    ),
    missing = "no"
  ) %>%
  bold_labels()


ny_ns %>%
  select(all_of(vars), Urban) %>%
  mutate(urb = case_when(
    Urban==1 ~ "Urban",
    Urban==0 ~ "Rural",
    TRUE~NA_character_
  )) %>%
  tbl_summary(
    by = urb,
    label = var_lab,
    type = all_continuous() ~ "continuous2", 
    statistic = all_continuous() ~ c(
      "{mean}",
      "{median} ({p25}, {p75})",
      "{min}, {max}"
    ),
    missing = "no"
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2)) %>%
  bold_labels()

```

